Appendix 4: Results

### 2016-02-22 Harris_2015/manuscript-materials/draft.Rmd
more manuscript and appendices

**Abstract:** Inferring species interactions from co-occurrence data is one of
the most controversial tasks in community ecology. One difficulty is that a
single pairwise interaction can ripple through an ecological network and produce
surprising indirect consequences. For example, the negative correlation between
two competing species can be reversed in the presence of a third species that 
outcompetes both of them. Here, I apply models from statistical physics, called 
Markov networks or Markov random fields, that can predict
the direct and indirect consequences of any possible species interaction matrix.
Interactions in these models can be estimated from observed co-occurrence
rates via maximum likelihood, controlling for indirect effects. Using simulated
landscapes with known interactions, I evaluated Markov networks and six existing
approaches. The Markov networks consistently outperformed other methods, 
correctly isolating direct interactions between species pairs even when indirect
interactions or abiotic factors largely overpowered them. Two computationally 
efficient approximations, based on controlling for indirect effects with linear 
or generalized linear models, also performed well. Null models showed no 
evidence of being able to control for indirect effects, and reliably yielded 
incorrect inferences when such effects were present.

@@

In the third set of landscapes, I simulated each species' abundance (instead of
just their presence/absence), and interactions between species occurred on a 
per-capita basis. To avoid runaway mutualisms leading to infinite abundance,
all interaction coefficients were negative in these simulations. Good 
performance on these landscapes would indicate some robustness to the
mechanistic details of species interactions.

@@

For the null model and the Markov network, the probability of rejecting the
null hypothesis of zero interaction was estimated across a range of "true"
interaction strengths using a generalized additive model with binomial responses
(reject versus not; Appendix 4). The Type I error rate was defined as this
model's estimated probability of rejecting the null hypothesis when the "true"
interaction was zero.

@@

Finally, we can evaluate the models' statistical inferences (focusing on the
first two simulation types, where the interaction coefficients are easiest to
interpret).  The Markov network's Type I error rate depended on the 
simulation type: [[0.02]] for simulations that matched the model's assumptions,
versus [[0.14]] for simulations that included environmental heterogeneity (see 
Appendix 4 for confidence interval coverage across a range of $\beta_{ij}$ 
values). The null model's Type I error rates were [[0.30]] and [[0.50]] for the
constant and heterogeneous landscapes, respectively---far higher than the 
nominal 0.05 rate. Figure 4C shows, across a range of true interaction 
strengths, the probability that the null model or the Markov network will 
predict the wrong sign of the interaction with 95% confidence.  The null model 
makes such errors more than [[5]] times as often as the Markov network, even
though it only rejects the null hypothesis [[twice]] as often overall (Appendix
4). The Markov network's errors were also more concentrated around 0, so it 
never misclassified strong interactions like the null model did.

>>>>

**Abstract:** Inferring species interactions from co-occurrence data is one of
the most controversial tasks in community ecology. One difficulty is that a
single pairwise interaction can ripple through an ecological network and produce
surprising indirect consequences. For example, the negative correlation between
two competing species can be reversed in the presence of a third species that 
outcompetes both of them. Here, I apply models from statistical physics, called 
Markov networks or Markov random fields, that can predict
the direct and indirect consequences of any possible species interaction matrix.
Interactions in these models can be estimated from observed co-occurrence
rates via maximum likelihood, controlling for indirect effects. Using simulated
landscapes with known interactions, I evaluated Markov networks and six existing
approaches. The Markov networks consistently outperformed other methods, 
correctly isolating direct interactions between species pairs even when indirect
interactions or abiotic factors largely overpowered them. Two computationally 
efficient approximations, based on controlling for indirect effects with partial
correlations or generalized linear models, also performed well. Null models 
showed no evidence of being able to control for indirect effects, and reliably 
yielded incorrect inferences when such effects were present.

@@

In the final set of landscapes, I simulated each species' abundance (instead of
just their presence/absence); here, interactions between species occurred on a 
per-capita basis. To prevent runaway mutualisms leading to infinite abundance,
all interaction coefficients were negative in these simulations. Good 
performance on these landscapes would indicate some robustness to the
mechanistic details of species interactions.

@@

For the null model and the Markov network, the probability of rejecting the
null hypothesis of zero interaction was estimated across a range of "true"
interaction strengths using a generalized additive model with binomial responses
(reject versus not; Appendix 4). The probability of rejection when the "true"
value of $\beta$ was zero was defined as the Type I error rate.

@@

Finally, we can evaluate the models' statistical inferences (focusing on the
first two simulation types, where the interaction coefficients are easiest to
interpret).  The Markov network's Type I error rate depended on the 
simulation type: 0.02 for simulations that matched the model's assumptions,
versus 0.14 for simulations that included environmental heterogeneity (see 
Appendix 4 for confidence interval coverage across a range of $\beta_{ij}$ 
values). The null model's Type I error rates were 0.31 and 0.53 for the
constant and heterogeneous landscapes, respectively---far higher than the 
nominal 0.05 rate. Figure 4C shows, across a range of true interaction 
strengths, the probability that the null model or the Markov network will 
predict the wrong sign of the interaction with 95% confidence.  The null model 
makes such errors more than 5 times as often as the Markov network, even
though it only rejects the null hypothesis twice as often overall (Appendix
4). The Markov network's errors were also more concentrated around 0, so it 
never misclassified strong interactions like the null model did (Figure 4C).

### 2016-02-22 Harris_2015/manuscript-materials/draft.tex
more manuscript and appendices

\textbf{Abstract:} Inferring species interactions from co-occurrence
data is one of the most controversial tasks in community ecology. One
difficulty is that a single pairwise interaction can ripple through an
ecological network and produce surprising indirect consequences. For
example, the negative correlation between two competing species can be
reversed in the presence of a third species that outcompetes both of
them. Here, I apply models from statistical physics, called Markov
networks or Markov random fields, that can predict the direct and
indirect consequences of any possible species interaction matrix.
Interactions in these models can be estimated from observed
co-occurrence rates via maximum likelihood, controlling for indirect
effects. Using simulated landscapes with known interactions, I evaluated
Markov networks and six existing approaches. The Markov networks
consistently outperformed other methods, correctly isolating direct
interactions between species pairs even when indirect interactions or
abiotic factors largely overpowered them. Two computationally efficient
approximations, based on controlling for indirect effects with linear or
generalized linear models, also performed well. Null models showed no
evidence of being able to control for indirect effects, and reliably
yielded incorrect inferences when such effects were present.

@@

When the number of species is larger than about 30, noralizing Equation
1 can become intractable, and researchers will either need to
approximate it (Lee and Hastie 2012) or approximate its gradient
({\textbf{???}}) if they want to fit a Markov network model. For the
analyses presented here, I used the rosalia package (Harris 2015a) for
the R programming language (R Core Team 2015) to calculate
\(p(\vec{y}; \alpha, \beta)\) and its gradients exactly (Murphy 2012);
the package passes these functions to the ``BFGS'' method in R's
general-purpose optimizer, which finds values for \(\alpha\) and
\(\beta\).

@@

For the null model and the Markov network, the probability of rejecting
the null hypothesis of zero interaction was estimated across a range of
``true'' interaction strengths using a generalized additive model with
binomial responses (reject versus not; Appendix 4). The Type I error
rate was defined as this model's estimated probability of rejecting the
null hypothesis when the ``true'' interaction was zero.

@@

Finally, we can evaluate the models' statistical inferences (focusing on
the first two simulation types, where the interaction coefficients are
easiest to interpret). The Markov network's Type I error rate depended
on the simulation type: {[}{[}0.02{]}{]} for simulations that matched
the model's assumptions, versus {[}{[}0.14{]}{]} for simulations that
included environmental heterogeneity (see Appendix 4 for confidence
interval coverage across a range of \(\beta_{ij}\) values). The null
model's Type I error rates were {[}{[}0.30{]}{]} and {[}{[}0.50{]}{]}
for the constant and heterogeneous landscapes, respectively---far higher
than the nominal 0.05 rate. Figure 4C shows, across a range of true
interaction strengths, the probability that the null model or the Markov
network will predict the wrong sign of the interaction with 95\%
confidence. The null model makes such errors more than {[}{[}5{]}{]}
times as often as the Markov network, even though it only rejects the
null hypothesis {[}{[}twice{]}{]} as often overall (Appendix 4). The
Markov network's errors were also more concentrated around 0, so it
never misclassified strong interactions like the null model did.

@@

Markov networks---particularly the Ising model for binary networks---are
very well understood, having been studied for nearly a century (Cipra
1987). Tapping into this framework would thus allow ecologists to take
advantage of into a vast set of existing discoveries and techniques for
dealing with indirect effects, stability, and alternative stable states.
Numerous extensions to the basic network are possible as well. For
example, the states of the interaction network can be modeled as a
function of the local abiotic environment (Lee and Hastie 2012,
{\textbf{???}}), which would lead to a better understanding of the
interplay between biotic and abiotic effects on community structure.
Alternatively, models could allow one species to alter the relationship
between two other species (Whittam and Siegel-Causey 1981, Tjelmeland
and Besag 1998, cf Bruno et al. 2003).

@@

>>>>

\textbf{Abstract:} Inferring species interactions from co-occurrence
data is one of the most controversial tasks in community ecology. One
difficulty is that a single pairwise interaction can ripple through an
ecological network and produce surprising indirect consequences. For
example, the negative correlation between two competing species can be
reversed in the presence of a third species that outcompetes both of
them. Here, I apply models from statistical physics, called Markov
networks or Markov random fields, that can predict the direct and
indirect consequences of any possible species interaction matrix.
Interactions in these models can be estimated from observed
co-occurrence rates via maximum likelihood, controlling for indirect
effects. Using simulated landscapes with known interactions, I evaluated
Markov networks and six existing approaches. The Markov networks
consistently outperformed other methods, correctly isolating direct
interactions between species pairs even when indirect interactions or
abiotic factors largely overpowered them. Two computationally efficient
approximations, based on controlling for indirect effects with partial
correlations or generalized linear models, also performed well. Null
models showed no evidence of being able to control for indirect effects,
and reliably yielded incorrect inferences when such effects were
present.

@@

When the number of species is larger than about 30, noralizing Equation
1 can become intractable, and researchers will either need to
approximate it (Lee and Hastie 2012) or approximate its gradient (Harris
2016) if they want to fit a Markov network model. For the analyses
presented here, I used the rosalia package (Harris 2015a) for the R
programming language (R Core Team 2015) to calculate
\(p(\vec{y}; \alpha, \beta)\) and its gradients exactly (Murphy 2012);
the package passes these functions to the ``BFGS'' method in R's
general-purpose optimizer, which finds values for \(\alpha\) and
\(\beta\).

@@

For the null model and the Markov network, the probability of rejecting
the null hypothesis of zero interaction was estimated across a range of
``true'' interaction strengths using a generalized additive model with
binomial responses (reject versus not; Appendix 4). The probability of
rejection when the ``true'' value of \(\beta\) was zero was defined as
the Type I error rate.

@@

Finally, we can evaluate the models' statistical inferences (focusing on
the first two simulation types, where the interaction coefficients are
easiest to interpret). The Markov network's Type I error rate depended
on the simulation type: 0.02 for simulations that matched the model's
assumptions, versus 0.14 for simulations that included environmental
heterogeneity (see Appendix 4 for confidence interval coverage across a
range of \(\beta_{ij}\) values). The null model's Type I error rates
were 0.31 and 0.53 for the constant and heterogeneous landscapes,
respectively---far higher than the nominal 0.05 rate. Figure 4C shows,
across a range of true interaction strengths, the probability that the
null model or the Markov network will predict the wrong sign of the
interaction with 95\% confidence. The null model makes such errors more
than 5 times as often as the Markov network, even though it only rejects
the null hypothesis twice as often overall (Appendix 4). The Markov
network's errors were also more concentrated around 0, so it never
misclassified strong interactions like the null model did (Figure 4C).

@@

Markov networks---particularly the Ising model for binary networks---are
very well understood, having been studied for nearly a century (Cipra
1987). Tapping into this framework would thus allow ecologists to take
advantage of into a vast set of existing discoveries and techniques for
dealing with indirect effects, stability, and alternative stable states.
Numerous extensions to the basic network are possible as well. For
example, the states of the interaction network can be modeled as a
function of the local abiotic environment (Lee and Hastie 2012, Harris
2016), which would lead to a better understanding of the interplay
between biotic and abiotic effects on community structure.
Alternatively, models could allow one species to alter the relationship
between two other species (Whittam and Siegel-Causey 1981, Tjelmeland
and Besag 1998, cf Bruno et al. 2003).

@@

\hyperdef{}{ref-harrisux5festimatingux5f2016}{\label{ref-harrisux5festimatingux5f2016}}
Harris, D. J. 2016. Estimating species interactions in large,
abiotically structured communities using markov networks and stochastic
approximation. bioRxiv.

### 2016-02-22 Harris_2015/show-results.Rmd
more manuscript and appendices

```{r, message = FALSE}
library(dplyr)
library(mgcv)
library(ggplot2)
library(tidyr)
library(knitr)
library(lme4)
```

## Import the results from Appendix 4:

@@

@@

#### Calculate average R-squared across methods and simulation types

@@

## Estimate R-squared uncertainty:

@@

@@

# Markov network is significant when lower bound is above zero or lower bound is
# below zero
markov_summary$sig_pos = markov_summary$lower > 0
markov_summary$sig_neg = markov_summary$upper < 0

confidently_right = function(data){
  (data$sig_pos & data$truth > 0) | (data$sig_neg & data$truth < 0)
}

confidently_wrong = function(data){
  (data$sig_pos & data$truth < 0) | (data$sig_neg & data$truth > 0)
}
```

## Create Figure 4:

@@

plotfun = function(..., add){
  if(add){
    lines(...)
  }else{
    plot(...)
  }
}

error_smoother = function(data, f){
  predict(
    gam(
      f(data) ~ s(truth),
      data = data,
      family = binomial
    ),
    data.frame(truth = truth_seq),
    type = "response"
  )
}

p_right_pairs = error_smoother(pairs_summary, confidently_right)
p_wrong_pairs = error_smoother(pairs_summary, confidently_wrong)

p_right_markov = error_smoother(markov_summary, confidently_right)
p_wrong_markov = error_smoother(markov_summary, confidently_wrong)

par(mfrow = c(1, 2))
plot(truth_seq, p_right_pairs + p_wrong_pairs, type = "l", ylim = c(0, 1), yaxs = "i")
lines(truth_seq, p_right_markov + p_wrong_markov, col = 2)
abline(h = 0.05)

plot(truth_seq, p_wrong_pairs, type = "l", ylim = c(0, 1), yaxs = "i")
lines(truth_seq, p_wrong_markov, col = 2)
par(mfrow = c(1, 1))

y_markov = error_smoother(markov_summary, confidently_wrong)

@@

@@

# R-squared for null versus correlation
round(summary(lm(null ~ I(correlation*sqrt(n_sites)), data = spread_estimates))$r.squared, 2)

@@

plotfun(
  truth_seq,
  y_pairs,
  type = "l",
  xlab = "\"True\" interaction strength",
  ylab = "P(confidently predict wrong sign)",
  bty = "l",
  yaxs = "i",
  col = 2,
  add = FALSE,
  ylim = c(0, .4),
  lwd = 2
)
mtext("C. Error rate vs.\ninteraction strength", side = 3, adj = 0, 
      font = 2, line = 1.2, cex = .9) 
plotfun(truth_seq, y_markov, add = TRUE, lwd = 2)
legend("topleft", lwd = 2, legend = c("Null model", "Markov network"), 
       col = c(2, 1), bty = "n")
dev.off()
```

@@

```{r}
# P(Pairs confidently wrong)
with(pairs_summary,  mean((sig_neg & truth > 0) | (sig_pos & truth < 0)))
# P(Markov network confidently wrong)
with(markov_summary, mean((sig_neg & truth > 0) | (sig_pos & truth < 0)))

# P(Pairs rejects null)
with(pairs_summary,  mean(sig_neg | sig_pos))

# P(Markov network rejects null)
with(markov_summary,  mean(sig_neg | sig_pos))

# P(reject null | small interaction) for Makrov network
# (approximate Type I error rates)
markov_summary %>%
  group_by(simulation_type) %>% 
  filter(abs(truth) < .1) %>% 
  summarize(Type_1_error_rate = mean(sig_neg | sig_pos)) %>%
  kable(digits = 3)

# P(reject null | small interaction) for Pairs
# (approximate Type I error rates)
pairs_summary %>%
  group_by(simulation_type) %>% 
  filter(abs(truth) < .1) %>% 
  summarize(Type_1_error_rate = mean(sig_neg | sig_pos)) %>%
  kable(digits = 3)

# Confidence interval coverage across all values
# What's the probability that any "true" value falls inside the 95% CI?
markov_summary %>% 
  group_by(simulation_type) %>% 
  summarize(mean(truth > lower & truth < upper)) %>%
  kable(digits = 3)
```

## Plot root mean square error versus "true" beta

```{r, cache = TRUE}
# Exclude the "abundance" simulations because the interpretation of $\beta$ 
# is different
rmses = markov_summary %>% 
  filter(simulation_type != "abund") %>% 
  mutate(rmse = sqrt((truth - estimate)^2))

ggplot(rmses, aes(x = truth, y = rmse)) + 
  geom_point(alpha = .2, size = 1) + 
  geom_smooth() + 
  theme_bw() + 
  coord_cartesian(ylim = c(0, max(rmses$rmse)))
```

## Plot confidence interval coverage versus true $\beta$ value

```{r, cache = TRUE}
# Coverage versus true beta value.  The top and bottom 0.5%
# of the distribution have been omitted to prevent bad behavior
# by the smoother in the tails.
markov_summary %>%
  filter(percent_rank(truth) > .005 & percent_rank(truth) < .995) %>%
  mutate(covered = truth > lower & truth < upper) %>%
  ggplot(aes(x = truth, y = as.integer(covered))) + 
  facet_grid(~simulation_type) + 
  geom_smooth(method = gam, formula = y ~ s(x), method.args = list(family = "binomial")) + 
  theme_bw() + 
  coord_cartesian(ylim = c(0, 1)) + 
  geom_vline(xintercept = 0) + 
  geom_hline(yintercept = 0.95, color = "red") + 
  ylab("Coverage")

>>>>

# A: Import packages and data

```{r, message = FALSE}
library(dplyr)
library(magrittr)
library(mgcv)
library(ggplot2)
library(tidyr)
library(knitr)
library(lme4)
```

## Import the results from Appendix 3:

@@

# B: Compare model estimates

@@

#### Average the R-squared values across methods and simulation types

@@

## Estimate uncertainty in mean R-squared:

@@

## Plot estimates versus "true" values across methods and lanscape sizes

The gray line indicates the 1:1 line, where models recover an unbiased estimate
of the "true" $\beta$ coefficients.  The GLM and Markov network methods estimate
the coefficients on the original scale, and approach unbiasedness as the number
of sites increases. The four methods based on (partial) correlations are
constrained to return values between -1 and 1, and do not return values on the
original scale.  The covariance and partial covariance (which are not scaled in
this way) may avoid this problem in future analyses (Loh and Wainwright 2013; 
see main text for full citation).

Note that the null model (Pairs) is not included in these graphs, as its
test statistic depends strongly on the size of the landscape and fills a range 
that would not fit on these graphs.

```{r, dev="png", dpi=500, cache = TRUE}
ggplot(filter(x, simulation_type == "no_env", method != "null"),
       aes(x = truth, y = estimate, color = factor(n_sites))) + 
  facet_wrap(~method) +
  geom_point(alpha = 0.01) +
  geom_abline(intercept = 0, slope = 1, size = 2, col = "darkgray") + 
  geom_smooth(method = "lm", fullrange = FALSE) + 
  coord_equal() + 
  scale_color_discrete(name = "n_sites")
```

# C: Inferential statistics

@@

# Markov network is significant when lower bound is above zero or lower bound is
# below zero
markov_summary$sig_pos = markov_summary$lower > 0
markov_summary$sig_neg = markov_summary$upper < 0
```

The `error_smoother` is a function that fits a generalized additive model to 
estimate the probability that a model estimate will match some criterion (e.g. 
statistical significance) for some parameter as a function of that parameter's
"true" value.  This function is used in Figure 4C and in calculating Type I
error rates below.

error_smoother = function(data, f, values = truth_seq){
  predict(
    gam(
      f(data) ~ s(truth),
      data = data,
      family = binomial
    ),
    data.frame(truth = values),
    type = "response"
  )
}
```

## Create Figure 4

Panels A and B plot the point estimates or test statistics for different models 
against one another.

Panel C shows the probability of rejecting the null hypothesis (of no 
interaction between two species) in the opposite direction of the true value
(see main text).

@@

@@

# Calculate p(confidently wrong) for each model as a function of
# the true values
confidently_wrong = function(data){
  (data$sig_pos & data$truth < 0) | (data$sig_neg & data$truth > 0)
}
y_markov = error_smoother(markov_summary, confidently_wrong)
y_pairs = error_smoother(pairs_summary, confidently_wrong)

@@

# R-squared for null versus correlation
round(
  summary(lm(null ~ I(correlation*sqrt(n_sites)), data = spread_estimates))$r.squared, 
  2
)

@@

plot(
  truth_seq,
  y_pairs,
  type = "l",
  xlab = "\"True\" interaction strength",
  ylab = "P(confidently predict wrong sign)",
  bty = "l",
  yaxs = "i",
  col = 2,
  ylim = c(0, .4),
  lwd = 2
)
mtext("C. Error rate vs.\ninteraction strength", side = 3, adj = 0, 
      font = 2, line = 1.2, cex = .9) 
lines(truth_seq, y_markov, lwd = 2)
legend("topleft", lwd = 2, legend = c("Null model", "Markov network"), 
       col = c(2, 1), bty = "n")
dev.off()
```

@@

#### P(Pairs confidently wrong):
```{r}
with(pairs_summary,  mean((sig_neg & truth > 0) | (sig_pos & truth < 0)))
```

#### P(Markov network confidently wrong)
```{r}
with(markov_summary, mean((sig_neg & truth > 0) | (sig_pos & truth < 0)))
```

#### P(Pairs rejects null)
```{r}
with(pairs_summary,  mean(sig_neg | sig_pos))
```

#### P(Markov network rejects null)
```{r}
with(markov_summary,  mean(sig_neg | sig_pos))
```

## Type I error rates:

Type I error rates (probability of rejecting the null hypothesis given a 
"true" interaction strength of zero, according to the spline model defined by
`error_smoother` above). The smoother focuses on the 
`r markov_summary %>% summarize(mean(abs(truth) < 1)) %>% round(., 2) %>% multiply_by(100)`%
of the data points whose "true" values fall between -1 and 1, because its 
estimates near 0 are more accurate when they aren't affected by distant data 
points.

#### Markov network Type I error rates
```{r}
markov_summary %>% 
  filter(abs(truth) < 1) %>% 
  group_by(simulation_type) %>% 
  do(data.frame(
    `Type I error rate` = 
      error_smoother(., function(x){x$sig_pos | x$sig_neg}, values = 0)
  )) %>%
  kable(digits = 3)
```

#### Null model Type I error rates
```{r}
pairs_summary %>% 
  filter(abs(truth) < 1) %>% 
  group_by(simulation_type) %>% 
  do(data.frame(
    `Type I error rate` = 
      error_smoother(., function(x){x$sig_pos | x$sig_neg}, values = 0)
  )) %>%
  kable(digits = 3)
```

## Plot Markov network confidence interval coverage versus true $\beta$ value

The top and bottom 0.5% of the distribution have been omitted to prevent bad 
behavior by the smoother in the tails. The horizontal red line shows the nominal
95% coverage rate and the black vertical line marks where the "true" value
of $\beta$ was zero. The Type I error rates calculated above for the Markov 
network closely match (one minus) the values of these curves at zero.

```{r}
coverage_data = markov_summary %>%
  filter(percent_rank(truth) > .005 & percent_rank(truth) < .995) %>%
  mutate(covered = truth > lower & truth < upper) %>%
  mutate(simulation_type = factor(simulation_type, c("no_env", "env", "abund")))

ggplot(coverage_data, aes(x = truth, y = as.integer(covered))) + 
  facet_grid(~simulation_type) + 
  geom_smooth(method = gam, formula = y ~ s(x), method.args = list(family = binomial)) + 
  theme_bw() + 
  coord_cartesian(ylim = c(0, 1), expand = FALSE) + 
  geom_vline(xintercept = 0) + 
  geom_hline(yintercept = 0.95, color = "red") + 
  ylab("Coverage")
